Management induced changes of soil organic carbon on global croplands

### 2020-11-09 SOCbudgetPaper/SOCbudget.Rmd
fix merging issues

---
title: Management induced changes of soil organic carbon on global croplands 
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Kristine
    surname: Karstens
    affiliation: 1
    email: kristine.karstens@pik-potsdam.de
    corresponding: true
  - given_name: Benjamin Leon
    surname: Bodirsky
    affiliation: 1
  - given_name: Your
    surname: Name
    affiliation: 2
  - given_name: Alexander
    surname: Popp
    affiliation: 1
affiliation:
  - code: 1
    address: Potsdam-Institut of Climate Impacts Research, Potsdam, Germany
  - code: 2
    address: Your affiliation 
abstract: |
  Soil organic carbon (SOC) is one of larges carbon stocks on Earth. It is three times larger than the vegetation pool and more than twice as big as the athmospheric pool, when looking into the first meter of the earth soil profile. Human cropping activties led and still lead to a depletion of SOC though, which are so far not well represented in global assessments of historic carbon emissions. While SOC models often represent well the biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated on the global scale. Here we create a spatial explicit data set for agricutural management on cropland, especially for crop residue and manure management, based on global historic production (FAOSTAT), and land-use (LUH2) data; and combine it with the IPCC Tier 2 approach to create a half degree resolution SOC stock changes on mineral soils. We estimate that due to arable farming soils have lost over 37 GtOC of which 4 GtOC have been regained within the period 1975-2010. We show that, our results on global scale based on Tier 2 IPCC methodolgy are in good agreement with Tier 1 default assumptions. We also find that SOC is very sensitive to management decision such as residue recycling indicating the nessessity to incorporated better management data in soil models.
bibliography: SOCbudget2.bib
running:
  title: R Markdown Template for Copernicus
  author: Nüst et al.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed.
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company.
availability:
  codedata: |
    Software code for paper and result prepartion can be found under www.github.com/k4rst3ns/. Data used for the output can be found under <something-with-zenodo>.  
authorcontribution: |
  Karstens wrote code and paper build on work of Bodirsky (and <here_could_be_your_names>). 
  Bodirsky, <here_could_be_your_names> and Popp revised paper.
disclaimer: |
  We like Copernicus.
acknowledgements: |
  Thanks to the rticles contributors!
output:
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
    md_extensions: +raw_attribute
  md_document:
    variant: markdown_github
  bookdown::word_document2: default
  rticles::copernicus_article: default
  bookdown::pdf_document2: default
---
\newpage

>>>>

---
title: Management induced changes of soil organic carbon on global croplands 
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Kristine
    surname: Karstens
    affiliation: 1
    email: kristine.karstens@pik-potsdam.de
    corresponding: true
  - given_name: Benjamin Leon
    surname: Bodirsky
    affiliation: 1
  - given_name: Your
    surname: Name
    affiliation: 2
  - given_name: Alexander
    surname: Popp
    affiliation: 1
affiliation:
  - code: 1
    address: Potsdam-Institut of Climate Impacts Research, Potsdam, Germany
  - code: 2
    address: Your affiliation 
abstract: |
  Soil organic carbon (SOC) is one of larges carbon stocks on Earth. It is three times larger than the vegetation pool and more than twice as big as the athmospheric pool, when looking into the first meter of the earth soil profile. Human cropping activties led and still lead to a depletion of SOC though, which are so far not well represented in global assessments of historic carbon emissions. While SOC models often represent well the biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated on the global scale. Here we create a spatial explicit data set for agricutural management on cropland, especially for crop residue and manure management, based on global historic production (FAOSTAT), and land-use (LUH2) data; and combine it with the IPCC Tier 2 approach to create a half degree resolution SOC stock changes on mineral soils. We estimate that due to arable farming soils have lost over 37 GtOC of which 4 GtOC have been regained within the period 1975-2010. We show that, our results on global scale based on Tier 2 IPCC methodolgy are in good agreement with Tier 1 default assumptions. We also find that SOC is very sensitive to management decision such as residue recycling indicating the nessessity to incorporated better management data in soil models.
bibliography: SOCbudget2.bib
running:
  title: R Markdown Template for Copernicus
  author: Nüst et al.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed.
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company.
availability:
  codedata: |
    Software code for paper and result prepartion can be found under www.github.com/k4rst3ns/. Data used for the output can be found under <something-with-zenodo>.  
authorcontribution: |
  Karstens wrote code and paper build on work of Bodirsky (and <here_could_be_your_names>). 
  Bodirsky, <here_could_be_your_names> and Popp revised paper.
disclaimer: |
  We like Copernicus.
acknowledgements: |
  Thanks to the rticles contributors!
output:
  bookdown::word_document2: default
  md_document:
    variant: markdown_github
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
    md_extensions: +raw_attribute
  rticles::copernicus_article: default
  bookdown::pdf_document2: default
---
\newpage

### 2020-11-09 SOCbudgetPaper/SOCbudget.tex
fix merging issues

\begin{equation}
\Delta SOC_{t, i} = SOC_{t, i} - SOC_{pnv, t, i}\quad, \qquad \quad  F^{SCF}_{t, i} = \frac{SOC_{t, i}}{SOC_{pnv, t, i}}.
\label{eq:totalstock}
\end{equation}

@@

According to \citep{sanderman_soil_2017} cumulative SOC emissions since the beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil with half of it attributed to grazing. It was also pointed out that these results might be conservatively low compared to experimental results. Considering the huge uncertainties in modelling SOC on the global scale, our estimate of 37 GtC in 2010 for cropland emissions only still seems consistent with \citep{sanderman_soil_2017} estimates.

Furthermore, the results of \citep{sanderman_soil_2017} modelled historical trends based on agricultural land expansion without considering SOC variations due to different management systems at all. \citep{pugh_simulated_2015} considered management effects like tillage and residue recycling in a static way, but neither changes over time nor alignment to observed historical data like yields-levels or no-tillage areas were taken into account. The study moreover concludes that productivity gains for crops while increasing yield levels by 18\%, do not spill over to a substantial decrease in SOC emissions, but only account for less than 1\% change in SOC emissions.

Our study for the first time uses a dynamic management dataset as driver for SOC dynamics. We show that the moderate global cropland expansion of around 11\% between 1974 and 2010 and the resulting depletion of SOC stocks in converted cropland has been out weighted by improved agricultural yields and practices. This is challenging \citep{pugh_simulated_2015} findings of only small effects due to improved practices.
Moreover, our sensitivity analysis indicates that (1) recycled residue biomass increases driven by higher yields as well as higher recycling rates, (2) improved manure recycling (e.g.~due to improved animal waste management system) and (3) the adoption of no tillage practices.

Modelling management effects on the global scale comes however with parametric and structural uncertainties. As pointed out by \citep{keel_large_2017} and \citep{smith_how_2020}, carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. \citep{keel_large_2017} questioned whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shift allometries and might be not reflected in outdated data sources. Following this argumentation SOC results shown in this study might overestimate actual SOC stocks due to an overestimation in residue biomass, which ratio compared to the harvested organ is normally reduced due to breeding. However, looking into management effect evaluation (see below) there is no indication for a systematic overrating of residue biomass, more likely we are still missing carbon inputs to the soil.

@@

To validate our modelled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors (\citet{ipcc_2006_2006}, \citet{ipcc_2019_2019}) which are based on measurement data for croplands (see \ref{tab:SCFtable}). To allow for comparison, we aggregated our stock change factors to the four IPCC climate zones.

\begin{figure}
\includegraphics[width=16cm]{../ResultNotebooks/Output/Images/TableSCF_comparison} \caption{This table shows different estimates of stock change factors $F^{SCF}$, defined as the SOC stock of managed croplands relative to the counterfactual SOC stock of undisturbed natural vegetation. We compare average values for four IPCC climate zone classifications. IPCC 2006 and 2019 default factors (medium input) as well as altered factors for low input of organic matter without any other subsystem consideration are compared to results for 1990 and 2010 of this study (SOC budget).}\label{fig:SCFtable}
\end{figure}

Our estimates correspond very well to the default stock-change factors used for the tier 1 estimation of IPCC, 2006. For the tropical regions the assumptions changed notably from the guidelines in 2006 to the update in 2019, leaving our results too low in comparison with IPCC, 2019. Considering yield gaps in mainly developing regions in the tropics the default assumption of medium input systems, might be an overestimation of actual SOC state. The additional effect of considering low input of biomass in tropical regions can however not explain the full mismatch to IPCC 2019 values but account for at least 5-7\% of it.

Looking again on the time trend our study shows the substantial impact of changed management on the development of \(F_{SCF}\).

@@

The world's SOC stock and its changes are highly uncertain (cite), which is visible by the wide range of global SOC stock estimates (see \ref{tab:SOCtable}).

@@

The global estimates of SOC stock by this study are on the lower end compared to other modelled results or more data driven estimates. Looking on regional results \ref{append:regcompare}, our estimates turn out to be in good agreement for most regions with the largest deviations for boreal areas. Considering that the model was parametrized for croplands, these mismatches are not surprising since the temperature effects on decomposition are fundamentally different for permafrost soils. To avoid that this bias influences our results, our study focuses exclusively on cropland soils, excluding most of the boreal zone. Moreover, when focussing on SOC changes, pristine natural vegetated areas without human land management under the same historic climatic conditions cancel out in the calculation of SOC emissions.

Our estimates for total SOC stocks of the world, as well as our SOC initialization are dominated by the representation of natural lands and pastures, which are however only estimated in a basic manner. For example, we do not have a differentiated parametrization of nitrogen and lignin content of litterfall for woody and grass type biomes. This leaves carbon inputs and decay behavior for natural land and pastures rather uncertain. The absolute values of stocks and emissions from land-use change therefore have to be used with caution. Especially in less forested areas the natural land representation might be off, due to parametrization assumptions of the natural litterfall.

Nevertheless, total SOC stocks are in a reasonable range and results on SOC stock changes in relative or absolute numbers are not altered by pristine vegetated areas at all. Moreover, the SOC gap from changed management is only affected from the natural vegetation representation in case of land use transitions. Then former natural land SOC stocks are declining to lower levels. However, as pointed out before, due to low cropland expansion rates, their SOC declines are outweighed by the general trend of increasing SOC stocks on existing cropland due to management improvements.
We conducted a sensitivity analysis for a wide range of possible parameter combinations for lignin and nitrogen parametrization of natural litterfall, which tend to change the global SOC stock substantially for different choices
(see appendix). It shows that the general trend of decreasing SOC gap is not altered even under very high estimates for natural SOC stocks.

>>>>

\begin{equation}
\Delta SOC_{t, i} = SOC_{t, i} - SOC_{pnv, t, i}\quad, \qquad \quad  F^{SCF}_{t, i} = \frac{SOC_{t, i}}{SOC_{pnv, t, i}}.
\label{eq:stockdiff}
\end{equation}

@@

According to \citep{sanderman_soil_2017} cumulative SOC emissions since the beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil with half of it attributed to grazing. Our estimate of 37 GtC in 2010 for cropland emissions only is twice as high as \citep{sanderman_soil_2017} estimations. However, there are large uncertainties in modelling SOC at the global scale, and the \citep{sanderman_soil_2017} results have been pointed out as potentially conservatively low compared with experimental results, leaving our results in an aceptable range.

Furthermore, \citep{sanderman_soil_2017} modelled historical trends based on agricultural land expansion without considering SOC variations due to different management systems at all. \citep{pugh_simulated_2015} considered management effects like tillage and residue recycling in a static way, but neither changes over time nor alignment to observed historical data like yields-levels or no-tillage areas were taken into account. The study moreover concludes that crop productivity gains (increasing yield levels by 18\%) do not lead to a~substantial decrease in SOC emissions, accounting for only less than 1\% change in SOC emissions.

Our study for the first time uses a dynamic management dataset as driver for SOC dynamics. We show that the moderate global cropland expansion of around 11\% between 1974 and 2010 and the resulting depletion of SOC stocks in converted cropland has been out weighted by improved agricultural yields and practices. This challenges \citep{pugh_simulated_2015} findings of only small effects due to improved practices.

Moreover, our sensitivity analysis indicates that (1) recycled residue biomass increases driven by higher yields as well as higher recycling rates play the most important role, followed by (2) improved manure recycling (e.g.~due to improved animal waste management systems) and (3) the adoption of no tillage practices.

Modelling management effects on the global scale comes however with parametric and structural uncertainties. As pointed out by \citep{keel_large_2017} and \citep{smith_how_2020}, carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. \citep{keel_large_2017} question whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shifts allometries which might not be reflected in outdated data sources. Following this line of argumentation, SOC results calculated in this study may overestimate actual SOC stocks due to an overestimation in residue biomass, for which the ratio compared to the harvested organ is normally reduced due to breeding. However, with an evaluation of management effects (see below), there is no indication for a systematic overrating of residue biomass. Rather, it is more likely that we are still missing carbon inputs to the soil.

@@

To validate our modelled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors (\citet{ipcc_2006_2006}, \citet{ipcc_2019_2019}) which are based on measurement data for croplands (see \ref{tab:SCFtable}). To allow for comparison, we aggregate our stock change factors to the four IPCC climate zones.

\begin{figure}
\includegraphics[width=16cm]{../ResultNotebooks/Output/Images/TableSCF_comparison} \caption{This table shows different estimates of stock change factors $F_{SCF}$, defined as the SOC stock of managed croplands relative to the counterfactual SOC stock of undisturbed natural vegetation. We compare average values for four IPCC climate zone classifications. IPCC 2006 and 2019 default factors (medium input) as well as altered factors for low input of organic matter without any other subsystem consideration are compared to results for 1990 and 2010 of this study (SOC budget).}\label{fig:SCFtable}
\end{figure}

Our estimates correspond very well to the default stock-change factors used for the tier 1 estimation of IPCC, 2006. For the tropical regions the assumptions changed notably from the guidelines in 2006 to the update in 2019, leaving our results too low in comparison with IPCC, 2019. Considering yield gaps in mainly developing regions in the tropics the default assumption of medium input systems might be an overestimation of actual SOC state. The additional effect of considering low inputs of biomass in tropical regions can however not explain the full mismatch to IPCC 2019 values but account for 14-21\% of it.

With regard to the time trend, our study shows the substantial impact of changing management factors on the development of \(F_{SCF}\).

@@

The world's SOC stock and its changes are highly uncertain (cite), which is seen in the wide range of global SOC stock estimates (see \ref{tab:SOCtable}).

@@

The global estimates of SOC stock from this study are on the lower end compared to other modelled results or more data driven estimates. Looking at regional results \ref{append:regcompare}, our estimates turn out to be in good agreement for most regions, with the largest deviations for boreal areas. Considering that the model was parametrized for croplands, these mismatches are not~suprising as the temperature effects on decomposition are fundamentally different for permafrost soils. To avoid that this bias influences our results, our study focuses exclusively on cropland soils, excluding most of the boreal zone. Moreover, when focussing on SOC changes on cropland, pristine natural vegetated areas without human land management under the same historic climatic conditions cancel out in the calculation of SOC emissions.

Our estimates for total SOC stocks of the world, as well as our SOC initialization, are dominated by the representation of natural lands and pastures, which are however only estimated in a basic manner. For example, we do not have a differentiated parametrization of nitrogen and lignin content of litterfall for woody and grass type biomes. This leaves carbon inputs and decay behavior for natural land and pastures rather uncertain. The absolute values of stocks and emissions from land-use change therefore have to be used with caution. Especially in less forested areas the natural land representation higher uncertainty may arise from the parametrization assumptions of natural litterfall.

Nevertheless, total SOC stocks are in a reasonable range and results for SOC stock changes in relative or absolute numbers are not altered by pristine vegetated areas at all. The SOC gap from changed management is only affected by the natural vegetation representation in the case of land use transitions. In this case, former natural land SOC stocks decline. However, as pointed out before, due to low cropland expansion rates, land use change-related SOC declines are outweighed by the general trend of increasing SOC stocks on existing cropland due to management improvements.
We conducted a sensitivity analysis for a wide range of possible parameter combinations for lignin and nitrogen parametrization of natural litterfall, which tend to change the global SOC stock substantially for different choices
(see appendix). This shows that the general trend of decreasing SOC gap is not altered even under very high estimates for natural SOC stocks.
